Assessing spatial correlations in pulsar timing arrays with the noise-marginalized optimal statistic

### 2018-03-19 manuscript/manuscript.tex
updates to manuscript

@@

This paper is organized as follows. In Sec.~\ref{sec:marg_os} 
we lay out the procedure for computing the noise-marginalized optimal statistic 
and compare to the fixed-noise optimal statistic. 
In Sec.~\ref{sec:spatial} determine how well 
the noise-marginalized optimal statistic can 
differentiate between different spatial correlations 
using simulated PTA data. 
In Sec.~\ref{sec:skyscrambles} we determine how well the noise-marginalized optimal statistic 
can assess the significance of a GW detection by performing sky scrambles \citep{cs2016}. 
We also compare our sky scramble results to the Bayesian analysis of sky scrambles done by \citet{tlb+2017}. 
We conclude in Sec.~\ref{sec:conclusion}.

@@

\sv{Justin, please bring notation up to date}
As introduced by \citet{abc+2009}, 
the optimal statistic is derived by analytically maximizing 
the PTA likelihood function in the weak-signal regime. 
We begin by finding the timing residuals for a given pulsar ${\bf{r}}_I$, which can be written as 
\begin{equation}
	{\bf{r}}_I = M {\bf{\epsilon}}_I + F {\bf{a}}_I + U {\bf{j}}_I + {\bf{n}}_I \,,\\
\end{equation}
where $M$ is the design matrix, 
$\epsilon_I$ is a vector of the timing model parameter uncertainties, 
${\bf{a}}_I$ are the red noise parameters describing both individual pulsars' red noise 
and common red noise in all pulsars, 
${\bf{j}}_I$ are white noise parameters describing frequency-correlated white noise, 
and ${\bf{n}}_I$ are white noise parameters describing uncorrelated white noise. 
The likelihood function is constructed from the residuals as given by
\begin{equation}
	p({\bf{r}}|\phi) = \frac{\exp[-\frac{1}{2} {\bf{r}}^T \Sigma^{-1}(\phi) {\bf{r}} ]}{\sqrt{\mathrm{det}[2\pi \Sigma(\phi)]}} \,,
\end{equation}
where $\bf{r}$ is an array of the timing residuals 
from the $M$ pulsars in the PTA,
\begin{equation}
	\bf{r} = \left[ \begin{array}{c} {\bf{r}}_1 \\ {\bf{r}}_2 \\ \vdots \\ {\bf{r}}_M \end{array} \right] \,,
\end{equation}
$\phi$ is the set of pulsar noise parameters, 
and $\Sigma$ is the covariance matrix of the residuals. 
We can write $\Sigma$ in terms of 
autocovariance and cross-covariance matrices ${\bf{P}}_I$ and ${\bf{S}}_{IJ}$, respectively, as
\begin{equation}
	\Sigma = \left[ \begin{array}{cccc} {\bf{P}}_1 & {\bf{S}}_{12} & \hdots & {\bf{S}}_{1M}  \\
							{\bf{S}}_{21} & {\bf{P}}_2 & \hdots & {\bf{S}}_{2M} \\
							\vdots & \vdots & \ddots & \vdots \\
							{\bf{S}}_{M1} & {\bf{S}}_{M2} & \hdots & {\bf{P}}_M \end{array} \right] \,,
\end{equation}
where
\begin{eqnarray}
	{\bf{P}}_I &=& \left\langle {\bf{r}}_I {\bf{r}}_I^T \right\rangle \,, \\
	{\bf{S}}_{IJ} &=& \left. \left\langle {\bf{r}}_I {\bf{r}}_J^T \right\rangle \right|_{I \neq J} \,.
\end{eqnarray}
In the frequency domain, these can be written as
\begin{eqnarray}
	\left\langle {\bf{r}}_I {\bf{r}}_I^T \right\rangle_{ij} &=& {\bf{R}}_I \left[ \int_{-\infty}^\infty df \; e^{2\pi i f \tau_{ij}} P_I(f) \right] {\bf{R}}_I^T \,, \\
	\left. \left\langle {\bf{r}}_I {\bf{r}}_J^T \right\rangle \right|_{ij} &=& \chi_{IJ} {\bf{R}}_I \left[ \int_{-\infty}^\infty df \; e^{2\pi i f \tau_{ij}} P_\mathrm{gw}(f) \right] {\bf{R}}_J^T \,,
\end{eqnarray}
where $i,j$ index the matrix elements, $\tau_{ij} \equiv |t_i - t_j|$ is the time interval between 
two observations, 
$\mathbf{R}$ is the timing model fitting matrix, 
$\chi_{IJ}$ is the overlap reduction function (ORF) that describes the spatial correlation between two pulsars' residuals, 
$P_\mathrm{gw}(f)$ is the GW power spectrum, 
and $P_I(f)$ is the power spectrum for the $I$th pulsar.

@@

When constructing the residuals, 
we typically fix the red noise parameters to the values that 
maximize the likelihood. 
However, this leads to a bias in the optimal statistic and SNR because 
the individual red noise parameters and the common red noise parameters 
are highly covariant. 
In this section, we compare three different techniques for computing the 
optimal statistic for PTAs with significant red noise. 
First, we fix the individual pulsars' red noise parameters to the 
maximum-likelihood values from individual pulsar analyses. 
Second, we fix the individual pulsars' red noise parameters to the 
values maximum-likelihood values from a common analysis 
involving all of the pulsars in our PTA. 
For the third method, we draw values of the individual pulsars' 
red noise parameters from the posteriors generated by a Bayesian analysis 
that includes both individual red noise and a common red noise process, 
allowing us to marginalize over the individual red noise.

We used these three techniques to compute the optimal statistic for simulated data sets. 
We constructed simulated data sets using the same procedure as in \citet{tlb+2017}. 
The simulated PTAs consisted of 10 MSPs with 
total observation times and noise properties chosen to be 
similar to the properties of the EPTA \cite{caballero+2016} (see Table~\ref{tab:sim}). 
We also injected a stochastic GWB amplitude with $\Agw = 5\times10^{-15}$.
\begin{table}[htb]
	\setlength{\tabcolsep}{5pt}
	\caption{Pulsar parameters used in simulated PTA}
	\begin{center}
	\begin{tabular}{ccccc}
		\hline\hline
    		Pulsar	& $T_\mathrm{obs}$ (yrs) & $\sigma_w$ ($\mu\mathrm{s}$) & $\log_{10}A_\mathrm{red}$ & $\gamma_\mathrm{red}$ \\
		\hline
		J0613$-$0200 & 16.054 & 1.58 & -13.90 & 3.18 \\
		J0751+1807 & 17.606 & 2.60 & -14.14 & 2.58 \\
		J1012+5307 & 16.831 & 1.47 & -13.09 & 1.65 \\
		J1640+2224 & 16.735 & 1.99 & -13.24 & 0.03 \\
		J1643$-$1224 & 17.300 & 1.65 & -18.56 & 4.04 \\
		J1713+0747 & 17.657 & 0.26 & -14.90 & 4.85 \\
		J1744$-$1134 & 17.250 & 0.65 & -13.60 & 2.00 \\
		J1857+0943 & 17.310 & 1.51 & -16.00 & 1.35 \\
		J1909$-$3744 & 9.379 & 0.12 & -13.99 & 2.06 \\
		J2145$-$0750 & 17.161 & 1.19 & -13.87 & 4.02 \\
    		\hline\hline
	\end{tabular}
	\end{center}
	\label{tab:sim}
\end{table}

The optimal statistic was computed using the \sv{fill in...} in PAL2 \CiteNeed. 
Figure~\ref{fig:os_dataset_sample} shows the fixed-noise and noise-marginalized 
optimal statistic for a single realization of the stochastic background. 
We also include the distribution of $\Agw^2$ found using the results of a Bayesian analysis 
for a common red process. For this particular realization of the stochastic background, 
the values of $\hat{A}^2$ from the noise-marginalized and Bayesian analyses are in good agreement 
with each other and the injected GW stochastic background, 
but the value from the fixed-noise analysis is significantly lower. 
The noise-marginalized analysis also gives a larger mean SNR compared to the fixed-noise analysis.
\begin{figure}[ht]
	\includegraphics[width=0.9\columnwidth]{plots/os_dataset50.pdf}
	\caption{Optimal statistic and SNR for a simulated dataset 
			containing an injected GW background with $\Agw = 5\times10^{-15}$. 
			We compare the values found using a fixed-noise analysis (dashed blue line) to the 
			mean optimal statistic and mean SNR found by marginalizing over 
			the individual pulsars' red noise parameters (solid orange line). 
			We also include the distribution of $\Agw^2$ found using the results of a Bayesian analysis 
			for a common red process (solid green line). 
			The dotted vertical line indicates the injected value, $\hat{A}^2 = 2.5 \times 10^{-29}$. 
			For this particular realization of the stochastic background, the fixed-noise analysis 
			gives $\hat{A}^2 = 1.9 \times 10^{-30}$ with $\mathrm{SNR} = 0.60$, 
			while the noise-marginalized analysis gives $\hat{A}^2 = 1.9\times10^{-29}$ with $\mathrm{SNR} = 1.1$. 
			For comparison, the Bayesian analysis gives a mean value of $\Agw^2 = 4.5\times10^{-29}$.}
	\label{fig:os_dataset_sample}
\end{figure}

>>>>

\usepackage{natbib}
\usepackage{amsmath,amssymb}

@@

This paper is organized as follows. In Sec.~\ref{sec:marg_os} 
we lay out the procedure for computing the noise-marginalized optimal statistic 
and compare to the fixed-noise optimal statistic. 
In Sec.~\ref{sec:spatial} determine how well 
the noise-marginalized optimal statistic can 
differentiate between different spatial correlations 
using simulated PTA data. 
In Sec.~\ref{sec:skyscrambles} we determine how well the noise-marginalized optimal statistic 
can assess the significance of a GW detection by performing sky scrambles \citep{cs2016}. 
%We also compare our sky scramble results to the Bayesian analysis of sky scrambles done by \citet{tlb+2017}. 
We conclude in Sec.~\ref{sec:conclusion}.

@@

\sv{bring notation up to date}
As introduced by \citet{abc+2009}, 
the optimal statistic is derived by analytically maximizing 
the PTA likelihood function in the weak-signal regime. 
We begin by finding the timing residuals for a given pulsar ${\bf{r}}_I$, which can be written as 
\begin{equation}
	{\bf{r}}_I = M {\bf{\epsilon}}_I + F {\bf{a}}_I + U {\bf{j}}_I + {\bf{n}}_I \,,\\
\end{equation}
where $M$ is the design matrix, 
$\epsilon_I$ is a vector of the timing model parameter uncertainties, 
${\bf{a}}_I$ are the red noise parameters describing both individual pulsars' red noise 
and common red noise in all pulsars, 
${\bf{j}}_I$ are white noise parameters describing frequency-correlated white noise, 
and ${\bf{n}}_I$ are white noise parameters describing uncorrelated white noise. 
The likelihood function is constructed from the residuals as given by
\begin{equation}
	p({\bf{r}}|\phi) = \frac{\exp[-\frac{1}{2} {\bf{r}}^T \Sigma^{-1}(\phi) {\bf{r}} ]}{\sqrt{\mathrm{det}[2\pi \Sigma(\phi)]}} \,,
\end{equation}
where $\bf{r}$ is an array of the timing residuals 
from the $M$ pulsars in the PTA,
\begin{equation}
	\bf{r} = \left[ \begin{array}{c} {\bf{r}}_1 \\ {\bf{r}}_2 \\ \vdots \\ {\bf{r}}_M \end{array} \right] \,,
\end{equation}
$\phi$ is the set of pulsar noise parameters, 
and $\Sigma$ is the covariance matrix of the residuals. 
We can write $\Sigma$ in terms of 
autocovariance and cross-covariance matrices ${\bf{P}}_I$ and ${\bf{S}}_{IJ}$, respectively, as
\begin{equation}
	\Sigma = \left[ \begin{array}{cccc} {\bf{P}}_1 & {\bf{S}}_{12} & \hdots & {\bf{S}}_{1M}  \\
							{\bf{S}}_{21} & {\bf{P}}_2 & \hdots & {\bf{S}}_{2M} \\
							\vdots & \vdots & \ddots & \vdots \\
							{\bf{S}}_{M1} & {\bf{S}}_{M2} & \hdots & {\bf{P}}_M \end{array} \right] \,,
\end{equation}
where
\begin{eqnarray}
	{\bf{P}}_I &=& \left\langle {\bf{r}}_I {\bf{r}}_I^T \right\rangle \,, \\
	{\bf{S}}_{IJ} &=& \left. \left\langle {\bf{r}}_I {\bf{r}}_J^T \right\rangle \right|_{I \neq J} \,.
\end{eqnarray}
In the frequency domain, these can be written as
\begin{eqnarray}
	\left\langle {\bf{r}}_I {\bf{r}}_I^T \right\rangle_{ij} &=& {\bf{R}}_I \left[ \int_{-\infty}^\infty df \; e^{2\pi i f \tau_{ij}} P_I(f) \right] {\bf{R}}_I^T \,, \\
	\left. \left\langle {\bf{r}}_I {\bf{r}}_J^T \right\rangle \right|_{ij} &=& \chi_{IJ} {\bf{R}}_I \left[ \int_{-\infty}^\infty df \; e^{2\pi i f \tau_{ij}} P_\mathrm{gw}(f) \right] {\bf{R}}_J^T \,,
\end{eqnarray}
where $i,j$ index the matrix elements, $\tau_{ij} \equiv |t_i - t_j|$ is the time interval between 
two observations, 
$\mathbf{R}$ is the timing model fitting matrix, 
$\chi_{IJ}$ is the overlap reduction function (ORF) that describes the spatial correlation between two pulsars' residuals, 
$P_\mathrm{gw}(f)$ is the GW power spectrum, 
and $P_I(f)$ is the power spectrum for the $I$th pulsar.

@@

When constructing the residuals, 
we typically fix the red noise parameters to the values that 
maximize the likelihood. 
However, this leads to a bias in the optimal statistic and SNR because 
the individual red noise parameters and the common red noise parameters 
are highly covariant. 
In this section, we compare three different techniques for computing the 
optimal statistic for PTAs with significant red noise. 
First, we fix the individual pulsars' red noise parameters to the 
maximum-likelihood values from individual pulsar analyses. 
Second, we fix the individual pulsars' red noise parameters to the 
values that jointly maximize the likelihood for a common analysis 
of all of the pulsars in our PTA 
that searches over the pulsars' red noise parameters and 
a common red process. 
For the third method, we draw values of the individual pulsars' 
red noise parameters from the posteriors generated by a Bayesian analysis 
that includes both individual red noise and a common red noise process, 
allowing us to marginalize over the individual red noise.

We used these three techniques to compute the optimal statistic for simulated data sets. 
We simulated PTA data sets with comprised of 18 MSPs 
with the same sky locations, observation times, and noise properties 
as the MSPs in the NANOGrav 9-year data set 
with at least 3 years of observation time \citep{abb+15} 
(see Table~\ref{tab:sim} for more details). 
To these data sets we injected an isotropic stochastic GWB 
\sv{insert strength of GWB... are we going to have one value or more than one?}.
\begin{table}[htb]
	\setlength{\tabcolsep}{5pt}
	\caption{Pulsar parameters used in simulated PTAs}
	\begin{center}
	\begin{tabular}{ccccc}
		\hline\hline
    		Pulsar	& $T_\mathrm{obs}$ (yrs) & $\sigma_w$ ($\mu\mathrm{s}$) & $\log_{10}A_\mathrm{red}$ & $\gamma_\mathrm{red}$ \\
		\hline
		J0030+0451 & & & \\
		J0613$-$0200 & 16.054 & 1.58 & -13.90 & 3.18 \\
		J1012+5307 & 16.831 & 1.47 & -13.09 & 1.65 \\
		J1024$-$0719 & & & \\
		J1455$-$3330 & & & \\
		J1600$-$3053 & & & \\
		J1614$-$2230 & & & \\
		J1640+2224 & 16.735 & 1.99 & -13.24 & 0.03 \\
		J1713+0747 & 17.657 & 0.26 & -14.90 & 4.85 \\
		J1741+1351 & & & \\
		J1744$-$1134 & 17.250 & 0.65 & -13.60 & 2.00 \\
		B1855+09 & & & \\
		J1853+1303 & & & \\
		J1909$-$3744 & 9.379 & 0.12 & -13.99 & 2.06 \\
		J1918$-$0642 & & & \\
		J2010$-$1323 & & & \\
		J2145$-$0750 & 17.161 & 1.19 & -13.87 & 4.02 \\
		J2317+1439 & & & \\
    		\hline\hline
	\end{tabular}
	\end{center}
	\label{tab:sim}
\end{table}

%We constructed simulated data sets using the same procedure as in \citet{tlb+2017}. 
%The simulated PTAs consisted of 10 MSPs with 
%total observation times and noise properties chosen to be 
%similar to the properties of the EPTA \cite{caballero+2016} (see Table~\ref{tab:sim}). 
%We also injected a stochastic GWB amplitude with $\Agw = 5\times10^{-15}$.
%\begin{table}[htb]
%	\setlength{\tabcolsep}{5pt}
%	\caption{Pulsar parameters used in simulated PTA}
%	\begin{center}
%	\begin{tabular}{ccccc}
%		\hline\hline
%    		Pulsar	& $T_\mathrm{obs}$ (yrs) & $\sigma_w$ ($\mu\mathrm{s}$) & $\log_{10}A_\mathrm{red}$ & $\gamma_\mathrm{red}$ \\
%		\hline
%		J0613$-$0200 & 16.054 & 1.58 & -13.90 & 3.18 \\
%		J0751+1807 & 17.606 & 2.60 & -14.14 & 2.58 \\
%		J1012+5307 & 16.831 & 1.47 & -13.09 & 1.65 \\
%		J1640+2224 & 16.735 & 1.99 & -13.24 & 0.03 \\
%		J1643$-$1224 & 17.300 & 1.65 & -18.56 & 4.04 \\
%		J1713+0747 & 17.657 & 0.26 & -14.90 & 4.85 \\
%		J1744$-$1134 & 17.250 & 0.65 & -13.60 & 2.00 \\
%		J1857+0943 & 17.310 & 1.51 & -16.00 & 1.35 \\
%		J1909$-$3744 & 9.379 & 0.12 & -13.99 & 2.06 \\
%		J2145$-$0750 & 17.161 & 1.19 & -13.87 & 4.02 \\
%    		\hline\hline
%	\end{tabular}
%	\end{center}
%	\label{tab:sim}
%\end{table}

The optimal statistic was computed using the PTA data analysis package 
PAL2 %\footnote{\href{https://github.com/jellis18/PAL2}{https://github.com/jellis18/PAL2}} 
\citep{evh17a}. 
Figure~\ref{fig:os_dataset_sample} shows the fixed-noise and noise-marginalized 
optimal statistic for a single realization of the stochastic background. 
We also include the distribution of $\Agw^2$ found using the results of a Bayesian analysis 
for a common red process. For this particular realization of the stochastic background, 
the values of $\hat{A}^2$ from the noise-marginalized and Bayesian analyses are in good agreement 
with each other and the injected GW stochastic background, 
but the value from the fixed-noise analysis is significantly lower. 
The noise-marginalized analysis also gives a larger mean SNR compared to the fixed-noise analysis.
\begin{figure}[ht]
	\includegraphics[width=0.9\columnwidth]{plots/os_dataset50.pdf}
	\caption{Optimal statistic and SNR for a simulated dataset 
			containing an injected GW background with $\Agw = 5\times10^{-15}$. 
			We compare the values found using a fixed-noise analysis (dashed blue line) to the 
			mean optimal statistic and mean SNR found by marginalizing over 
			the individual pulsars' red noise parameters (solid orange line). 
			We also include the distribution of $\Agw^2$ found using the results of a Bayesian analysis 
			for a common red process (solid green line). 
			The dotted vertical line indicates the injected value, $\hat{A}^2 = 2.5 \times 10^{-29}$. 
			For this particular realization of the stochastic background, the fixed-noise analysis 
			gives $\hat{A}^2 = 1.9 \times 10^{-30}$ with $\mathrm{SNR} = 0.60$, 
			while the noise-marginalized analysis gives $\hat{A}^2 = 1.9\times10^{-29}$ with $\mathrm{SNR} = 1.1$. 
			For comparison, the Bayesian analysis gives a mean value of $\Agw^2 = 4.5\times10^{-29}$.}
	\label{fig:os_dataset_sample}
\end{figure}
